---
title: "Class 18: Pertussis mini-project"
author: "Georgina Canto-Encalada"
format: pdf
---

Pertusssis case numbers are tracked by the CDC and made available here: https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

> Q1.  With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time

We will use the datapasta pacjage to "scrape" thus data.

```{r}
cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
         No..Reported.Pertussis.Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )
```

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

Using ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)
ggplot(cdc, aes(x=Year,y=No..Reported.Pertussis.Cases)) + geom_point() +geom_line() + geom_vline(xintercept=1946, col="blue") + geom_vline(xintercept=1996, col="red")  + geom_vline(xintercept = 2020,col="yellow")


```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

ANS: The number of people rejecting the vaccination increased

# Read data from the CMI-PB project

It returns JSON data that we can read into R with the help of a package called **jsonlite**

```{r}
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE)
head(subject)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```


```{r}
specimen<-read_json("http://cmi-pb.org/api/specimen",simplifyVector = TRUE)
head(specimen)
```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?



```{r}

library(lubridate)
library(dplyr)


subject$age <- today() - ymd(subject$year_of_birth)

```


The average age of aP individuals

```{r}
# aP group
ap <- subject %>% filter(infancy_vac == "aP")
round( summary( time_length( ap$age, "years" ) ) )
```

The average age of wP individuals

```{r}
# wP group
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

are they significantly different?

ANS: Yes

```{r}
# Statistical test to compare average ages
# Assuming normal distribution and equal variances
t_test_result <- t.test(age ~ infancy_vac, data = subject)

# Print the t-test result
print(t_test_result)

```

> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
boost_age <- time_length(int, "year")
head(boost_age)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

Yes, they are different.

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

# Joining multiple tables

Let's join subject and specimen list based on subject_id. We will use the `inner_join()` from dplyr package

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details


```{r}

meta <- inner_join(specimen,subject)
head(meta)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

Another inner_join to add all the meta data onto these measurements

```{r}
ab_titer <- read_json("http://cmi-pb.org/api/v4/plasma_ab_titer",simplifyVector = TRUE)
abdata <- inner_join(ab_titer,meta)
head(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

How many Ab isotyopes are there?

```{r}
table(abdata$isotype)
```


>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

ANS: More samples were collected during 2020 than in 2022. It has sense since it was the COVID outbreak.

```{r}
table(abdata$dataset)
```



> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
ggplot(igg) + 
  aes(MFI_normalised,antigen) +
  xlim(0,75) +
  geom_boxplot() +
  facet_wrap(vars(visit), nrow=2)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

ANS: FIM2/3. This antigen is associated with Bordetella pertussis infection. 


```{r}
ggplot(abdata) + 
  aes(MFI,antigen, col=infancy_vac) +
  geom_boxplot() 
  
```
> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = T) +
  facet_wrap(vars(visit)) +
  theme_bw()
```
```{r}
filter(igg, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = T) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

> Q16. What do you notice about these two antigens time courses and the PT data in particular?

It seems that there is a distinction between the immune responses to different vaccine formulations across different visits. The PT antigen, which is a component of the pertussis vaccine, appears to induce a variable immune response over time, which is evident from the spread of the data points.

> Q17. Do you see any clear difference in aP vs. wP responses?

No

```{r}
#pt_data<-filter(abdata,antigen=="PT",isotype=="IgG", dataset == "2020_dataset")
pt_data<-filter(abdata,antigen=="PT",isotype=="IgG",dataset=="2021_dataset")
```

```{r}
table(pt_data$dataset)

```

```{r}

ggplot(pt_data) +
  aes(x=actual_day_relative_to_boost, y=MFI_normalised, col=infancy_vac, 
      group=subject_id) + 
  geom_point() +
  geom_line()
  

```

> Q18. Does this trend look similar for the 2020 dataset?

No

```{r}
pt_data<-filter(abdata,antigen=="PT",isotype=="IgG", dataset == "2020_dataset")
ggplot(pt_data) +
  aes(x=actual_day_relative_to_boost, y=MFI_normalised, col=infancy_vac, 
      group=subject_id) + 
  geom_point() +
  geom_line()

```












